---
title: "Cyclistics"
author: "omar moharrem"
date: "2023-04-03"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```





# import and install tools

```{r importing and installing tools}
library(tidyverse)
library(dplyr)
library(lubridate)
library(magrittr)
library(ggplot2)
library(geosphere)
install.packages("snakecase", repos = "http://cran.us.r-project.org")
library(snakecase)
```
# import data
```{r}
 may <- read.csv("202004-divvy-tripdata.csv")
jun <- read.csv("202005-divvy-tripdata.csv")
jul <- read.csv("202006-divvy-tripdata.csv")
aug <- read.csv("202007-divvy-tripdata.csv")
sep <- read.csv("202008-divvy-tripdata.csv")
oct <- read.csv("202009-divvy-tripdata.csv")
nov <- read.csv("202010-divvy-tripdata.csv")
dec <- read.csv("202011-divvy-tripdata.csv")
jen_21 <-read.csv("202012-divvy-tripdata.csv")
feb_21 <- read.csv("202101-divvy-tripdata.csv")
mar_21 <- read.csv("202102-divvy-tripdata.csv")
apr_21 <- read.csv("202103-divvy-tripdata.csv")
```
## aggregate data
```{r}
aggregate_data_frame <- rbind(may, jun, jul, aug, sep, oct, nov, dec,jen_21,feb_21,mar_21,apr_21)
```
## familiarize with data
```{r}
head(aggregate_data_frame)
summary(aggregate_data_frame)
```
# data cleaning
```{r}
clean_data <- na.omit(aggregate_data_frame)
```
### check
```{r}
head(clean_data)
colSums(is.na(clean_data))
```
### the starting time must be before the ending time so in this section i cleared the rows which starting time is after the ending time
```{r}
clean_data <- clean_data[clean_data$started_at <= clean_data$ended_at, ]
```
## check
```{r}
clean_data %>%
  filter(started_at > ended_at)
```
## dealing with duplicate rows
```{r}
clean_data <- subset(clean_data, !duplicated(clean_data))
```
### check
```{r}
which(duplicated(clean_data))
```
# calculation,metrics and grouping 
## ride duration calculation
```{r}
clean_data$ride_duration_min <- as.numeric(difftime(clean_data$ended_at, clean_data$started_at, units = "mins"))
head(clean_data)
```
## changing of number of casual and member customers among the time
### Create a new data frame with the count of casual and member riders
```{r}
riders_by_month <- clean_data %>%
  group_by(year = lubridate::year(started_at), month = lubridate::month(started_at, label = TRUE)) %>%
  summarize(casual_riders = sum(member_casual == "casual"),
            member_riders = sum(member_casual == "member"))
```
## group the data by month and rider type and calculate the number of rides
```{r}
rides_by_month_and_rider <- clean_data %>% 
  group_by(month = lubridate::month(started_at), member_casual) %>% 
  summarize(total_rides = n())
```
## group by month and year and count the number of rides
```{r}
# group by month and year and count the number of rides
rides_per_month_year <- clean_data %>%
  group_by(year = lubridate::year(started_at), month = lubridate::month(started_at, label = TRUE)) %>%
  summarize(total_rides = n()) %>%
  arrange(year, month)
```
## create a new column for hour of day
```{r}
clean_data$hour <- lubridate::hour(clean_data$started_at)
```

## calculate distance between start and end coordinates
### Convert longitude and latitude values to radians
```{r}
lat1 <- clean_data$start_lat * pi/180
lon1 <- clean_data$start_lng * pi/180
lat2 <- clean_data$end_lat * pi/180
lon2 <- clean_data$end_lng * pi/180
```
## Calculate distance between the two points using the Vincenty formula
```{r}
clean_data$distance_meters <- distVincentyEllipsoid(cbind(lon1, lat1), cbind(lon2, lat2))
```
## weekdays column
```{r}
clean_data$day_of_week <- weekdays(as.Date(clean_data$started_at))
clean_data <- clean_data %>% 
  mutate(day_order = recode(day_of_week, 
                            "Monday" = 1, 
                            "Tuesday" = 2, 
                            "Wednesday" = 3, 
                            "Thursday" = 4, 
                            "Friday" = 5, 
                            "Saturday" = 6, 
                            "Sunday" = 7))
```
## comparing casual and member rides in the week days
```{r}
daily_rides <- clean_data %>%
  group_by(day_of_week, member_casual) %>%
  summarize(total_rides = n()) %>%
  ungroup()
```
## create a new data frame with the counts of ride trips and ride duration starting from each station
```{r}
start_stations <- clean_data %>% 
  group_by(start_station_name) %>% 
  summarize(total_trips = n(), ride_duration_min = sum(ride_duration_min)) %>% 
  arrange(desc(total_trips))
start_stations <- start_stations %>% filter(total_trips != 39218)
```
## top 10 stations in ride duration
```{r}
start_stations_top10_ride_duration <- start_stations %>%
  top_n(10,ride_duration_min )
```
## top 10 stations in total trips
```{r}
start_stations_top10_total_trips <- start_stations %>%
  top_n(10,total_trips )
```

# statistical brief
```{r}
print(mean(clean_data$ride_duration_min))
print(max(clean_data$ride_duration_min))
mode_weekday <- mode(clean_data$day_order)
cat("The mode weekday is", mode_weekday, "\n")
```


# data vizzes
## compare riding duration of casual and member riders
```{r}
ggplot(clean_data, aes(x = member_casual, y = ride_duration_min)) + 
  geom_bar(stat = "summary") +
  labs(title = "Ride Duration by Rider Type", x = "Rider Type", y = "Ride Duration (minutes)")
```
## comparing number of casual and member trip rides in each month
```{r}
ggplot(rides_by_month_and_rider, aes(x = month, y = total_rides/1000, fill = member_casual)) + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Monthly Rides by Rider Type", x = "Month", y = "Number of Rides/1000") +
  scale_x_discrete(labels = c( "May", "June","July", "August", "September", "October", "November", "December" , "January", "February", "March", "April"))
```
## Create a line chart of the count of casual and member riders by month
```{r}
ggplot(data = riders_by_month, aes(x = month, y = casual_riders/1000, group = year, color = "Casual Riders")) +
  geom_line() +
  geom_point() +
  geom_line(aes(y = member_riders/1000, color = "Member Riders")) +
  geom_point(aes(y = member_riders/1000, color = "Member Riders")) +
  labs(title = "Number of Casual and Member Riders by Month",
       x = "Month",
       y = "Number of Riders/1000",
       color = "Member Type") +
  scale_color_manual(values = c("Member Riders" = "blue", "Casual Riders" = "red"))
```
## create a bar chart with rider type on the x-axis and hour of day on the y-axis
```{r}
ggplot(clean_data, aes(x = member_casual, fill = factor(hour))) +
  geom_bar(position = "dodge") +
  scale_fill_discrete(name = "Hour of Day", labels = c("Midnight", "", "2 AM", "", "4 AM", "", "6 AM", "", "8 AM", "", "10 AM", "", "Noon", "", "2 PM", "", "4 PM", "", "6 PM", "", "8 PM", "", "10 PM")) +
  labs(title = "Number of Trips by Hour of Day and Rider Type",
       x = "Rider Type",
       y = "Number of Trips") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 14, face = "bold"),
        axis.title.y = element_text(size = 14, face = "bold"),
        plot.title = element_text(size = 16, face = "bold"))
```
## comparing Casual and member distance
```{r}
ggplot(clean_data, aes(x = member_casual, y = distance_meters)) +
  geom_bar(stat = "summary", fun = "median", fill = "#4e79a7") +
  labs(title = "Distance Covered by Rider Type",
       x = "Rider Type",
       y = "Distance (M)")
```
## comparing casual and member rides in the week days
```{r}
ggplot(daily_rides, aes(x = day_of_week, y = total_rides/1000, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Number of Rides by Day of Week and Rider Type", x = "Day of Week", y = "Total Rides/1000") +
  scale_fill_manual(values = c("#0072B2", "#F0E442"), labels = c("Member", "Casual")) +
  theme_minimal()
```
## plot the top 10 stations in total trips
```{r}
ggplot(start_stations_top10_total_trips, aes(x = start_station_name, y = total_trips)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Number of Ride Trips Starting from Each Station", x = "Start Station", y = "Number of Trips")
```
## plot the top 10 stations in ride duration
```{r}
ggplot(start_stations_top10_ride_duration, aes(x = start_station_name, y = ride_duration_min)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = " Total Ride duration from Each Station", x = "Start Station", y = "Ride duration min")
```

